A Wait-free Queue with Polylogarithmic Step Complexity

### 2022-10-19 thesis/thesis.tex
finalize draft

\begin{abstract}
In this work, we introduce a novel linearizable wait-free queue implementation.
Linearizability and lock-freedom are standard requirements for designing shared data structures. To the best of our knowledge, all of the existing linearizable lock-free queues in the literature have a common problem in their worst case called the CAS Retry Problem. We show our algorithm resolves this problem and has worst case running time better than prior lock-free queues. The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue} in our algorithm, is $O(\log^2 p + \log q)$,  where $p$ is the number of processes and $q$ is the size of the queue when the operation is linearized.
\end{abstract}

@@

A system is called \it{asynchronous} when processes in the system run at arbitrary varying speeds, i.e., the scheduling of each process is independent from the scheduling of other processes. Our model is an asynchronous shared-memory distributed system of $p$ processes. Herlihy showed \cite{10.1145/114005.102808} that we one cannot implement concurrent version of all data structures with multi-reader multi-writer registers alone, introducing a hieachy on how powerful are objects to solve consensus problem among processes. Objects like LL/SC, CAS can be used to reach consensus among any number of processors. CAS object implements an atomic Compare \& Swap(\nf{new, old}) operation, which if the value stored in the object is \nf{old}, it updates the value to \nf{new} and returns {true}, otherwise returns \nf{false}.  In this work we use Compare \& Swap (\nf{CAS}) objects to reach universal synchronization.

@@

We will present a wait-free linearizable queue, which to the best of our knowledge  is the first wait-free bounded poly-log time queue. We design a tournament tree in which processes try to propagate their operations up to the root to be linearized and they can do queries to get information about the linearization stored in the root. To solve the CAS Retry problem when a process is propagating its operation to a node we create a helping mechanism that ensures operations are propagated after at most 2 \nf{CAS} invocations. The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue}, is $O(\log^2 p + \log q)$,  where $q$ is the size of the queue when the operation is linearized. \nf{CAS} instructions cost more than other instructions for processor. Each operation in our queue does at most $14\log p$ \nf{CAS} operations. We use unbounded memory space in our queue, we present a way to reduce the memory size to the total number of operations. The garbage collection remains open which we have pointed to a way we think it can be handled.

@@

\begin{figure}[hbt]
  \center\includegraphics[scale=0.3]{pics/baskets}
  \caption{Baskets queue idea. There is a time that all operations in a basket were running concurrently, but only one has succeeded to do \texttt{CAS}. To order the operations in a basket, the mechanism in the algorithm for processes is to \texttt{CAS} again. The successful process will be the next one in the basket and so on.}
\end{figure}
\bf{TODO: make circles the same size}

Ladan-Mozes and Shavit~\cite{DBLP:journals/dc/Ladan-MozesS08} presented an optimistic approach to implement a queue. MS-queue uses two \nf{CAS}es to do an enqueue: one to change the tail to the new node and another one to change the next pointer of the previous added node to the new node. They use a doubly-linked list to do fewer \texttt{CAS} operations in an \nf{Enqueue} than MS-queue. Asin  previous algorithms, the worst case happens in the case where the contention is high when $p$ concurrent enqueues happene, their nodes have to be appended to the linked list one by one. The amortized worst-case complexity is still $\Omega(p)$ \texttt{CAS}es.

@@

Gidenstam, Sundell, and Tsigas~\cite{DBLP:conf/opodis/GidenstamST10} introduced a new algorithm using a linked list of arrays. The queue is stored in a shared array where head and tail pointers point to the current elements in the queue. When the array is occupied an empty array is linked to the array and tail pointers are updated. Global head points to the array containing the first element in the queue, and head that points to the first element in that array. Global tail and tail pointers are similar (see~Figure~\ref{fig::sundell}). The pointers might be behind their real places, a process updates the position of the pointers when using them. When a process wants to write an element to current head it uses \nf{CAS}, so if $p$ processes try to enqueue simultaneously, the step (also \nf{CAS}) complexity remains $\Omega(p)$. Their design is lock-free but not wait-free.

@@

Milman et al.~\cite{DBLP:conf/spaa/MilmanKLLP18} designed a lock-free queue supprting futures. A future is an operation that can be applied on the queue when it is requested. Their idea of batching allows a sequence of operations to be submitted as a batch for later execution on the MS-queue. They use some properties of the size of the queue before and after a batch which is similar to a part of our work. Their queue is not wait-free, in fact if the batch sizes are 1 then the queue is like MS-queue and it supports a safety condition weaker than linearizabilty introduced by the authors, called Extended Medium Futures Linearizability.

Nikolaev and Ravindran~\cite{DBLP:journals/corr/abs-2201-02179} present a wait-free queue that uses the fast-path slow-path methodology. Their work is based on a circular queue using bounded memory. They design a slow phase which is wait-free bounded. If a process has no progress other processes try to help it in its \nf{CAS} loop in its slow phase. Since there can be $p$ processes helping, the worst-case complexity cannot be better than $\Omega(p)$.

@@

\subsection{Universal Constructions and other Poly-log Time Data Structures}
Herlihy discussed the possibility of implementing shared objects from other objects~\cite{10.1145/114005.102808}. A \textit{universal construction} is an algorithm that can implement a shared version of any given sequential object. We can implement a concurrent queue using a universal construction. Jayanti proved an $\Omega(\log p)$ lower bound on the worst-case shared-access time complexity of $p$-process universal constructions~\cite{DBLP:conf/podc/Jayanti98a}. He also mentions that universal construction by Afek, Dauber, and Touitou~\cite{DBLP:conf/stoc/AfekDT95} can be modified to $O(\log n)$ worst case step complexity, using atomic access to $\Omega(n \log n)$ bit words. Chandra, Jayanti and Tan  introduced a semi-universal construction that achieves $\textsc{O}(\log^2 p)$ shared accesses~\cite{DBLP:conf/podc/ChandraJT98}. Their universal construction cannot be used to create any data structure, but a set of objects called closed objects. We mention a non-practical universal construction with poly-log \nf{CAS} invocations in page 12.

Ellen and Woelfel introduced an implementation of a Fetch\&Inc object with step complexity of $O(\log p)$ using $O(\log n$)-bit \texttt{LL/SC} objects, where $n$ is the number of operations~\cite{10.1007/978-3-642-41527-2_20}. Their idea to achieve logarithmic complexity is to use a tree storing the Fetch\&Incs added by processes and when a process wants to do Fetch\&Inc it adds a Fetch\&Inc to the tree and return the number of elements in the tree. If we have an array and two Fetch\&Inc objects head and tail we can implement a queue: to do an enqueue we can Fetch\&Inc head and write in the value fetched of the array and to do a dequeue we can Fetch\&Inc the tail and fetched the array's cell in the value returned.

\subsection{Attiya--Fouren Lower Bound}
Because of the \nf{CAS} retry problem in previous list based queues one might guess $p$ factor is inherent in time complexity of concurrent queues. Attiya and Fouren proved a lower bound on amortized complexity of lock-free queues from the factor of contention (number of concurrent processes). Their result says any implementation of queues, using reads, writes and conditional operations like \nf{CAS} has $O(\sqrt{\log \log p})$ amortized complexity \cite{DBLP:conf/opodis/AttiyaF17}. The surprising point is that their result does not apply on our design and we manage to reach poly-log time complexity. It is an open problem to create the reduce the gap between our algorithm and Attiya--Fouren lower bound.

@@

>>>>

\begin{abstract}
In this work, we introduce a novel linearizable wait-free queue implementation.
Linearizability and lock-freedom are standard requirements for designing shared data structures. To the best of our knowledge, all of the existing linearizable lock-free queues in the literature have a common problem in their worst case called the CAS Retry Problem. We show our algorithm avoids this problem and has worst case running time better than prior lock-free queues. The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue} in our algorithm, is $O(\log^2 p + \log q)$,  where $p$ is the number of processes and $q$ is the size of the queue when the operation is linearized.
\end{abstract}

@@

A system is called \it{asynchronous} when processes in the system run at arbitrarily varying speeds, i.e., the scheduling of each process is independent from the scheduling of other processes. Our model is an asynchronous shared-memory distributed system of $p$ processes. Herlihy showed \cite{10.1145/114005.102808} that one cannot implement concurrent versions of all data structures with multi-reader multi-writer registers alone and introduced a hierarchy on how powerful objects are to solve the consensus problem among processes. Objects like LL/SC, CAS can be used to reach consensus among any number of processors. A CAS object provides an atomic Compare \& Swap(\nf{new, old}) operation: if the value stored in the object is \nf{old}, it updates the value to \nf{new} and returns \nf{true}, otherwise it returns \nf{false}.  In this work we use Compare \& Swap (\nf{CAS}) objects to synchronize among  processes.

@@

We will present a wait-free linearizable queue, which to the best of our knowledge  is the first wait-free queue whose operations run in a poly-logarithmic number of steps. We design a tournament tree in which each process tries to propagate its operations along a path from the process's leaf up to the root to be linearized. Processes can do queries to get information about the linearization stored in the root. To avoid the CAS Retry problem, when a process is propagating its operation to a node we create a helping mechanism that ensures operations are propagated after at most 2 \nf{CAS} invocations. The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue}, is $O(\log^2 p + \log q)$,  where $q$ is the size of the queue when the operation is linearized. \nf{CAS} instructions cost more than other instructions for processor. Each operation in our queue does $\Theta(\log p)$ \nf{CAS} operations compared to $\Omega(p)$ \nf{CAS} steps for previous queues in the worst case. We use unbounded memory space in our queue, we present a way to reduce the memory size to the total number of operations. We do not address garbage collection but we describe a way we think it can be handled.

@@

\begin{figure}[hbt]
  \center\includegraphics[scale=0.3]{pics/baskets}
  \caption{Baskets queue idea. There is a time that all operations in a basket were running concurrently, but only one has succeeded to do \texttt{CAS}. To order the operations in a basket, the mechanism in the algorithm for processes is to \texttt{CAS} again. The successful process will be the next one in the basket and so on.}
\end{figure}
% \bf{TODO: make small circles the same size}

Ladan-Mozes and Shavit~\cite{DBLP:journals/dc/Ladan-MozesS08} presented an optimistic approach to implement a queue. MS-queue uses two \nf{CAS}es to do an enqueue: one to change the tail to the new node and another one to change the next pointer of the previous added node to the new node. They use a doubly-linked list to do fewer \texttt{CAS} operations in an \nf{Enqueue} than MS-queue. As in  previous algorithms, the worst case happens in the case where the contention is high: when $p$ concurrent enqueues happen, their nodes have to be appended to the linked list one by one. The amortized worst-case complexity is still $\Omega(p)$ \texttt{CAS}es.

@@

Gidenstam, Sundell, and Tsigas~\cite{DBLP:conf/opodis/GidenstamST10} introduced a new algorithm using a linked list of arrays. The queue is stored in a shared array where head and tail pointers point to the current elements in the queue. When the array is occupied an empty array is linked to the array and tail pointers are updated. A global head points to the array containing the first element in the queue, and each process has a local head index that points to the first element in that array. Global tail and local tail pointers are similar (see~Figure~\ref{fig::sundell}). A process updates the position of the pointers after it does an operation. One process might go to sleep before setting the pointers, so the pointers might be behind their real places. They mention how to scan the arrays to update pointers while doing an operation. A process writes an element in the location head by a \nf{CAS} instruction, so if $p$ processes try to enqueue simultaneously, the step (and \nf{CAS}) complexity remains $\Omega(p)$. Their design is lock-free but not wait-free.

@@

Milman et al.~\cite{DBLP:conf/spaa/MilmanKLLP18} designed a lock-free queue supporting futures. In their queue, operations  return future objects instead of responses. Later when the response is needed it can be evaluated from the future object. They also define a weaker linearizabilty condition such that each operation can be linearized between its invocation and when its future is evaluated. Their idea of batching allows a sequence of operations to be submitted as a batch for later execution on the MS-queue. They use some properties of the size of the queue before and after a batch which is similar to a part of our work. Their queue is not wait-free, in fact if the batch sizes are 1 then the queue is like MS-queue.

Nikolaev and Ravindran~\cite{DBLP:journals/corr/abs-2201-02179} present a wait-free queue that uses the fast-path slow-path methodology. Their work is based on a circular queue using bounded memory. When a process wishes to do an enqueue or a dequeue, it starts two paths. One is faster but its running time is not bounded the other one is slower but has a time bound. They show that these two paths do not effect each other and the queue remains consistent. If a process has no progress other processes help its slow path to finish. The helping phase suffers from CAS Retry Problem, because processes compete in a \nf{CAS} loop to decide which succeeds to help. Because of this, the worst-case complexity cannot be better than $\Omega(p)$.

@@

\subsection{Universal Constructions and other Poly-log Time Data Structures}
A \textit{universal construction} is an algorithm that can implement a shared version of any given sequential object, introduced by Herlihy~\cite{10.1145/114005.102808}. We can implement a concurrent queue using a universal construction. Jayanti proved an $\Omega(\log p)$ lower bound on the worst-case shared-access time complexity of $p$-process universal constructions~\cite{DBLP:conf/podc/Jayanti98a}. He also mentions that the universal construction by Afek, Dauber, and Touitou~\cite{DBLP:conf/stoc/AfekDT95} can be modified to $O(\log p)$ worst case step complexity, using atomic access to $\Omega(p \log p)$ bit words. Chandra, Jayanti and Tan  introduced a semi-universal construction that achieves $\textsc{O}(\log^2 p)$ shared accesses using reasonably sized words~\cite{DBLP:conf/podc/ChandraJT98}. Their universal construction cannot be used to create any data structure, but a set of objects called closed objects (a queue is not a closed object). We mention a non-practical universal construction with a poly-log number of \nf{CAS} instructions on page 13.

% TODO: page

Ellen and Woelfel introduced an implementation of a Fetch\&Inc object with step complexity of $O(\log p)$ using $O(\log n$)-bit \texttt{LL/SC} objects, where $n$ is the number of operations~\cite{10.1007/978-3-642-41527-2_20}. Their idea to achieve logarithmic complexity is to use a tree storing the Fetch\&Incs added by processes. When a process wants to do Fetch\&Inc it adds a Fetch\&Inc to the tree and return the number of elements in the tree. There are  some similarities between designing a queue and a Fetch\&Inc object. A Fetch\&Inc object can be constructed from a queue. The algorithm by Ellen and Woelfel is interesting because it is the first wait-free data structure achieving poly-log implementation in a while.

\subsection{Attiya--Fouren Lower Bound}
Because of the \nf{CAS} retry problem in previous list-based queues one might guess the $\Omega(p)$ term is inherent in time complexity of concurrent queues. Attiya and Fouren gave a lower bound on amortized complexity of lock-free queues with regard ti $c$, the number of concurrent processes. Their result says if $c$ is $O(\log\log p)$, any implementation of queues using reads, writes and conditional operations like \nf{CAS} has $O(\sqrt{\log \log p})$ amortized complexity \cite{DBLP:conf/opodis/AttiyaF17}. The surprising point is that their result does not contradict ours and we manage to reach poly-log time complexity.

@@

\subsection{Attiya--Fouren Lower Bound}
As we talked earlier in Section 2.4 Attiya--Fouren lower bound says a concurrent implementation of queues using reads, writes and conditional operations like \nf{CAS} has $O(\sqrt{\log \log p})$ amortized complexity \cite{DBLP:conf/opodis/AttiyaF17} when the number of concurrent processes is $O(\log\log p)$. Our amortized complexity is $O(log^2p+log q)$. It is an open problem to the reduce the gap between our algorithm and Attiya--Fouren lower bound.
